Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE)

### 2014-02-15 WSSSPE1_report/WSSSPE1_summary_report.tex
Section 3.1: Keynote 1: Good, better, best, never let it rest.

\subsection{A recipe for sustainable software, Philip E. Bourne.}

@@

Bourne's keynote presentation was entitled ``A Recipe for Sustainable
Software.''  He emphasized that sustainability for software ``does not
just mean more money from Government.''  He noted that in places,
development in science has improved thanks to open source and
GitHub~\cite{github-web}, but for the most part remains arcane. He
argued that there is much we can learn from the App Store model. He
mentioned BioJava~\cite{biojava-web} and Open Science Data
Cloud~\cite{osdc-web} as distribution channels.  On a related note,
Bourne observed a common evolutionary pathway for computational
biology projects, from data archive to analytics platform to
educational use, and suggested that use of scientific software for
outreach might be the final step.

Bourne shared with the audience a few real challenges he
encountered. First, he has looked into reproducibility in
computational biology, and has concluded that ``I have proved I cannot
reproduce research from my own lab.''~\cite{Veretnik}

Another problem he experienced was staff retention with respect to
private organizations which reward those combining research and
software expertise (the ``Google Bus''). However, he is a strong fan
of private sector sustainability sources and public-private
partnerships. He noted that making a successful business from
scientific software alone is very rare: founders overvalue while
customers undervalue. He noted that to last, an open source project
needs a minimal funding requirement even with a vibrant community --
goodwill only goes so far if you are being paid to do something else.
He talked about grant schemes of relevance in the U.S., particularly
with regard to technology transfer~\cite{sbir-web, fased-web}.

He also had problems with selling research software: the technology
transfer office in his university wanted huge IP reach through,
whereby they would get a share of profits from drugs developed by
pharmaceutical companies who use the software.  He was aware this was
unrealistic but the technology transfer office insisted for a
while. He wants to push a one-click approach to purchasing
university-written software.

He then presented the usual arguments on directly valuing software as
a research output alongside papers. Readers will be familiar with this
debate. An interesting reference here is~\cite{peer-review-code}.

On the idea of the \emph{digital enterprise}, where IT underpins the
whole of organizational activities, he contended that universities are
way behind the curve on this. In particular, he highlighted the
separation of research, teaching and administration into silos without
a common IT framework as a blocker to many useful organizational
innovations: ``University 2.0 is yet to happen.'' He spoke of a
circumstance where someone had used an algorithm developed for
computational biology in marketing.

The role of an institution is important in this space. He argued that
funders can help train institutions not just individuals.

He concluded with a reference to his paper~\cite{bourne_ten} and
argued that computational scientists ``have a responsibility to
convince their institutions and communities that software is
scholarship,'' a point with which we strongly agree.

>>>>

\subsection{A Recipe for Sustainable Software, Philip E. Bourne.}

@@

Bourne's keynote presentation was entitled ``A Recipe for Sustainable
Software,'' developed based on his experiences.  He emphasized that
sustainability for software ``does not just mean more money from
Government'' (see also Section~\ref{sec:cross-cutting}).  Other
factors to consider, he mentioned, include costs of production, ease
of maintainence, community involvement, and distribution channels.

In places, Bourne said, development in science has improved thanks to
open source and GitHub~\cite{github-web}, but for the most part
remains arcane. He argued that we can learn much from the App Store
model about interfaces, ratings, and so on. He also mentioned
BioJava~\cite{biojava-web} and Open Science Data Cloud~\cite{osdc-web}
as distribution channels.  On a related note, Bourne observed a common
evolutionary pathway for computational biology projects, from data
archive to analytics platform to educational use, and suggested that
use of scientific software for outreach might be the final step.

Bourne shared with the audience a few real challenges he
encountered. First, also an anecdote, he has looked into
reproducibility in computational biology, but has concluded that ``I
have proved I cannot reproduce research from my own
lab.''~\cite{Veretnik}

Another problem he experienced was staff retention with respect to
private organizations which reward those combining research and
software expertise (the ``Google Bus''). However, he is a strong fan
of software sustainability through public-private partnerships. He
noted that making a successful business from scientific software alone
is very rare: founders overvalue while customers undervalue. He noted
that to last, an open source project needs a minimal funding
requirement even with a vibrant community --- goodwill only goes so
far if one is being paid to do something else.  He talked about grant
schemes of relevance in the U.S., particularly with regard to
technology transfer~\cite{sbir-web, fased-web}.

He also had problems with selling research software: the technology
transfer office in his university wanted huge intellectural property
reach through, whereby they would get a share of profits from drugs
developed by pharmaceutical companies who use the software.  He was
aware this was unrealistic but the technology transfer office insisted
for a while. He wants to push a one-click approach for customers to
purchase university-written software.

He then presented the usual arguments on directly valuing software as
a research output alongside papers. Readers will be familiar with this
debate.  An interesting reference provided by him
is~\cite{peer-review-code}, which explores involving software
engineers in the review process of scientific code.

On the notion of \emph{digital enterprise}, where information
technology (IT) underpins the whole of organizational activities, he
contended that universities are way behind the curve. In
particular, he highlighted the separation of research, teaching, and
administration into silos without a common IT framework as a blocker
to many useful organizational innovations: ``University 2.0 is yet to
happen.'' He spoke of a circumstance where someone had used an
algorithm developed for computational biology in marketing.
The role of an institution is important in this space. He argued that
funders can help train institutions, not just individuals, in this regard.

He concluded with a reference to his paper~\cite{bourne_ten} and
argued that computational scientists ``have a responsibility to
convince their institutions, reviewers, and communities that software is
scholarship, frequently
more valuable than a research article,'' a point with which we strongly agree.
